---
title: "ANOVA"
---

We will calculate power and sample size for a one-way ANOVA, a two-way ANOVA, and an ANCOVA. We will use base R, as well as the `pwr`, `pwr2`, `powertools` and `Superpower` packages. 

```{r, warning=FALSE, message=FALSE}
library(pwr)
library(pwr2)
library(powertools)
library(Superpower)
```

## One-way ANOVA

We use a one-way ANOVA to calculate power and sample size when we want to get the difference in means across groups that are from a single factor. 

For power estimation based on sample size we will use: 

* Number of groups and their means (or the group effects)
* Group sample sizes 
* Population standard deviation 

Imagine we have a book club with three groups that each use a different medium to interact with the book. Group A reads the book, group b listens to the audiobook, and group c watches the movie. We want to know how well each group understood the story line. Each group has 20 people, the expected population group means are 8, 5, and 12 and the populated standard deviation is expected to be 7. 

We will calculate the power  with $\alpha$ = 0.05. 

Using base R:

```{r}
power.anova.test(groups = 3,      # 3 groups
                 between.var = 13,# between group variance
                 within.var = 49, # within group variance
                 sig.level=0.05,  # sig level 
                 n = 20)          # no. people in each group 
```

Using the package powertools, we call `anova1way.F.bal()` to calculate the overall sample size for balanced groups (each group has 20 people). 

```{r}
anova1way.F.bal(n = 20, mvec = c(8, 5, 12), sd = 7, power = NULL) |> 
  round(2)
```

We can also calculate sample size with the `pwr` package. Here, we need to provide the effect size which is specified by the argument `f`. 

Cohen’s f is an effect size measure used in tests that evaluate significant differences among group means, such as one or two way ANOVA's. Cohen's f effect size defines defines effect sizes as: f = 0.1 as small, f = 0.25 as medium, and f = 0.4 as large, effect sizes (Cohen, 1988). 

We will first calculate Cohen's f by hand: 

```{r}
# Calculate Cohen's f 
book.means <- c(8, 5, 12)
sd   <- 7

k <- length(book.means)
mu <- mean(book.means)

# Calculate between group var 
between.var.pop <- sum((book.means - mu)^2) / k 
within.var <- sd^2 # within group var 

f <- sqrt(between.var.pop / within.var)

f
```

Now, we use the function `es.anova.f()` from the package `powertools` to calculate f and provide the group means and standard deviation. We get the same result as above. 

```{r}
# Use powertools to get the effect size 
f.pt <- es.anova.f(means = book.means, sd = sd, v = F)

f.pt
```
We now calculate power using the function `pwr.anova.test()` with our `f` and `f.pt`. 

```{r}
# Power analysis using f from manual calculation 
pwr.anova.test(k = 3, f = f, n = 20, sig.level = 0.05)

# Power analysis using f calculated by powertools
pwr.anova.test(k = 3, f = f.pt, n = 20, sig.level = 0.05)
```

For base R, `powertools`, and `pwr`, we see that our power estimate is 0.8 telling us that for the given parameters we specified, we have an 80% chance of correctly rejecting the null hypothesis

Now, we will calculate sample size for a one-way ANOVA when we know the power we want to achieve, but we need to determine the sample size.

What you need: 

* Number of groups 
* Group means or effect size 
* Significance level 
* Power

Using base R, we leave n (sample size) empty and input the power we would like to achieve. We will include a power of 0.8, again indicating that if a difference exists between the mean and target, we have an 80% chance of detecting it.

```{r}
power.anova.test(groups = 3, 
                 between.var = 13,
                 within.var = 49,
                 sig.level = 0.05,
                 power = 0.8) 
```

Using `powertools`:

```{r}
anova1way.F.bal(n = NULL, mvec = c(8, 5, 12), sd = 7, power = 0.8)
```

Our estimated sample size is 20 people per group.  

Using the `pwr` package and inputting an effect size of 0.2 instead of the estimated mean values: 

```{r}
# Get f using powertools
f.2way <- es.anova.f(means = c(8, 5, 12), sd = 7, v = F)

pwr.anova.test(k = 3, # groups
               f = f.2way, # effect size 
               sig.level = 0.05, power = 0.8)
```

Again, we see that our estimated sample size is 20. 

## Two-way ANOVA

We will run examples of power and sample size for a two-way ANOVA, which is used when you want to compare the difference in means across two factors. 

What you need

* Number of factor levels
* Group means and standard deviation 
* Significance level 
* Target power 

In our example, we want to evaluate how fertilizer and soil type influence plant growth. We have two types of fertilizer and two types of soil. Our dependent variable is plant growth. 

Before setting up the full study, we wanted to make sure it is sufficiently powered, so we ran a pilot study and found the mean values. We want to calculate statistical power of 0.9 at a 95% confidence level.  

```{r}
# Generate matrix of mean plant values 
plant.mat <- matrix(c(15.2, 19.1, 11.9, 13.3), nrow = 2, byrow = TRUE)

# Print  
plant.mat
```

Using the `powertools` package:

```{r}
anova2way.F.bal(n = NULL, mmatrix = plant.mat, sd = 3, alpha = 0.05,
power = 0.9, v = F)
```

We see the sample sizes required to detect the main effect of Factor A, B, and their interaction. 

Using the `pwr2` package, we can calculate the power estimate. First, need to calculate Cohen's f effect sizes. We can get these values for a balanced two-way ANOVA using the `powertools` package and inputting the plant matrix and the estimated standard deviation:

```{r}
# Get Cohen's f effect sizes using powertools
cd.anova2way <- es.anova.f(means = plant.mat, sd = 3, v = FALSE)

print(cd.anova2way)

```

The effect sizes for factors A, B, and their interaction(fAB) are returned. We will input fA and fB into our sample size calculation 

```{r}
pwr.2way(a = 2, b = 2, # n factors 
         alpha = 0.05, # significance level 
         size.A = 10, size.B = 10, # sample size 
         f.A = cd.anova2way["fA"], f.B = cd.anova2way["fB"]) # effect size
```

The output shows that there is a high chance of detecting an effect for Factor A (power.A ~ 0.99), and a relatively moderate chance of detecting an effect for Factor B (power.B ~ .78). Overall, the design is well powered for detecting effects of Factor A but less so for Factor B. The `power = 0.78` result is the lowest power returned among the factors tested. 

Using the `pwr` package, we can calculate the estimated sample size for a two-way ANOVA 

```{r}
ss.2way(a = 2, b = 2, #2 groups in factors A and B
        alpha = 0.05, #significance level 
        beta = 0.1, #Using 90% power
        f.A = 0.2, f.B = 0.3, #Effect sizes for factors A and B
        B = 100) #Number of iterations
```
With the input parameters, our sample size per group is 67, or 268 total. 

## ANCOVA

With an ANCOVA, we extend our one-way ANOVA by including a variable that we want to account for in the model. 

What we need: 

* Group means and standard deviation 
* Estimated $R^2$ 
* Significance level 
* Power 

When looking at story comprehension based on the medium (book, audiobook, or movie), we want to account for age. We will use the function `anova1way.F.bal()` from the `powertools` package and input our mean values, standard deviation, $R^2$, the number of covariates, and the power.  

```{r}
anova1way.F.bal(n = NULL, 
                mvec = c(8, 5, 12),
                sd = 7, 
                Rsq = 0.6^2, #estimated r^2 
                ncov = 1, #n covariates 
                power = 0.8)
```

Using the `Superpower` package 

```{r}
power_oneway_ancova(n = NULL,
                    mu = c(8, 5, 12),
                    sd = 7,
                    n_cov = 1, 
                    r2 = 0.6^2, 
                    beta_level = 0.8,
                    alpha_level = 0.05)
```


Similarly, we could compute the power by leaving the argument `beta_level = NULL` and inputting the estimated sample sizes per group: 

```{r}
power_oneway_ancova(n = c(15, 15, 15),
                    mu = c(25, 25, 32),
                    sd = 7,
                    n_cov = 1, 
                    r2 = 0.4^2, 
                    beta_level = NULL,
                    alpha_level = 0.05)
```


References 

*  Cohen J. Statistical Power Analysis for the Behavioral Sciences.
Lawrence Erlbaum Associates, Hillsdale, New Jersey, 2nd edition, 1988.