---
title: "ANOVA"
---

We will calculate power and sample size for a one-way ANOVA, a two-way ANOVA, and an ANCOVA. We will use base R, and the pwr and powertools packages. 

```{r, warning=FALSE, message=FALSE}
library(pwr)
library(pwr2)
library(powertools)
```

## One-way ANOVA

We use a one-way ANOVA to calculate power and sample size when we want to get the difference in means across groups that are from a single factor. 

For power estimation based on sample size we will use: 

* Number of groups and their means (or the group effects)
* Group sample sizes 
* Population standard deviation 

Imagine we have a book club with three groups that each use a different medium to interact with the book. Group A reads the book, group b listens to the audiobook, and group c watches the movie. We want to know how well each group understood the story line. Each group has 20 people, the expected population group means are 8, 5, and 12 and the populated standard deviation is expected to be 7. 

We will calculate the power  with $\alpha$ = 0.05. 

Using base R:

```{r}
power.anova.test(groups = 3,      # 3 groups
                 between.var = 13,# between group variance
                 within.var = 49, # within group variance
                 sig.level=0.05,  # sig level 
                 n = 20)          # no. people in each group 
```

Using powertools, we use `anova1way.F.bal()` to calculate the overall sample size for balanced groups (each group has 20 people) 

```{r}
anova1way.F.bal(n = 20, mvec = c(8, 5, 12), sd = 7, power = NULL) |> 
  round(2)
```

Using pwr, we need to provide the effect size. We will first calculate it by hand: 

```{r}
# Calculate Cohen's f 
book.means <- c(8, 5, 12)
sd   <- 7

k <- length(book.means)
mu <- mean(book.means)

# Calculate between group var 
between.var.pop <- sum((book.means - mu)^2) / k 
within.var <- sd^2 # within group var 

f <- sqrt(between.var.pop / within.var)

f
```

Now, we use the function `es.anova.f` from powertools to calculate f. We get the same result as above 

```{r}
# Use powertools to get the effect size 
f.pt <- es.anova.f(means = book.means, sd = sd, v = F)

f.pt
```

```{r}
# Power analysis using f from manual calculation 
pwr.anova.test(k = 3, f = f, n = 20, sig.level = 0.05)

# Power analysis using f calculated by powertools
pwr.anova.test(k = 3, f = f.pt, n = 20, sig.level = 0.05)
```

For base R, powertools, and pwr, we see that our power estimate is 0.8. 

Now, we will calculate sample size for a one-way ANOVA when we know the power we want to achieve, but we need to determine the sample size.

What you need: 

* Number of groups 
* Group means or effect size 
* Significance level 
* Power

Using base R, we leave n (sample size) empty and input the power we would like to achieve. We will include a power of 0.8, indicating that if a difference actually exists between the mean and target, we have a 80% chance of detecting it.

```{r}
power.anova.test(groups = 3, 
                 between.var = 13,
                 within.var = 49,
                 sig.level = 0.05,
                 power = 0.8) # We input the power as 0.8
```

Using powertools:

```{r}
anova1way.F.bal(n = NULL, mvec = c(8, 5, 12), sd = 7, power = 0.8)
```

Our estimated sample size is 21 per group.  

Using pwr and inputting an effect size of 0.2 instead of estimated mean values.  

```{r}
# Get f using {powertools}
f.2way <- es.anova.f(means = c(8, 5, 12), sd = 7, v = F)

pwr.anova.test(k = 3, # groups
               f = f.2way, # effect size 
               sig.level = 0.05, power = 0.8)
```

Our estimated sample size is about 21. 

## Two-way ANOVA

Example of power and sample size for a two-way ANOVA, which is used when you want to compare the difference in means across two factors 

What you need

* Number of factor levels
* Group means and standard deviation 
* Significance level 
* Target power 

In our example, we want to evaluate how fertilizer and soil type influence plant growth. We have two types of fertilizer and 2 types of soil. Our dependent variable is plant growth. 

Before setting up the full study, we wanted to make sure it is sufficiently powered so we ran a pilot study and found the mean values. We want to calculate statistical power of 0.9 at a 95% confidence level.  

```{r}
# Generate matrix of mean values 
plant.mat <- matrix(c(15.2, 19.1, 11.9, 13.3), nrow = 2, byrow = TRUE)

# Print matrix to view 
plant.mat
```

Using powertools:

```{r}
anova2way.F.bal(n = NULL, mmatrix = plant.mat, sd = 3, alpha = 0.05,
power = 0.9, v = F)
```

Using pwr2 for a balanced two-way ANOVA 

```{r}
# Get Cohen's f effect sizes using powertools
cd.anova2way <- es.anova.f(means = plant.mat, sd = 3, v = FALSE)

print(cd.anova2way)

```

The effect sizes for factors A, B, and their interaction(fAB) are returned. We will input fA and fB into our sample size calculation 

```{r}
pwr.2way(a = 2, b = 2, # n factors 
         alpha = 0.05, # significance level 
         size.A = 10, size.B = 10, # sample size 
         f.A = cd.anova2way["fA"], f.B = cd.anova2way["fB"]) # effect size
```


## ANCOVA

With an ANCOVA, we extend our one-way ANOVA by including a variable that we want to account for in the model. 

What we need: 

* Group means and standard deviation 
* Estimated $R^2$ 
* Significance level 
* Power 

When looking at story comprehension based on the medium (book, audiobook, movie), we want to account for age. We will use the function `anova1way.F.bal` from powertools and input our mean values, standard deviation, $R^2$, the number of covariates, and the power.  

```{r}
anova1way.F.bal(n = NULL, 
                mvec = c(8, 5, 12),
                sd = 7, 
                Rsq = 0.6^2, #estimated r^2 
                ncov = 1, #n covariates 
                power = 0.8)
```

