---
title: "Regression"
---

## Multiple Regression

## Overall F test

For sample size estimation based on power, we need the following:

-   hypothesized effect size (f^2^ = R^2^ / (1 − R^2^))
-   numerator degrees of freedom (u = number of predictors being tested, occasionally represented as 'p')
-   significance level of test
-   power of test

For power estimation based on sample size, we need the following:

-   hypothesized effect size (f^2^ = R^2^ / (1 − R^2^))
-   numerator degrees of freedom (u = number of predictors being tested, occasionally represented as 'p')
-   denominator degrees of freedom (v = n − u − 1, where n is sample size). Hence, n = v + u +1
-   significance level of test

### Example: sample size

Suppose there is an annual 5k running race in a city. We are interested if we can predict an individual's time to complete the race using two variables: total hours spent training the past three months and average pace per mile during training runs. We can perform a multiple regression to model race completion time as a function of training time and pace. We want to sample enough subjects to detect an effect when our two independent (explanatory) variables explains 45% of variation of the dependent (response) variable, race completion time. Assume the following:

-   hypothesized effect size of (f^2^ = .45 / (1-.45)), f^2^ = 0.82; R^2^ = .45

-   numerator degrees of freedom u = 2 (two predictors)

-   significance level = 0.01

-   desired power = 0.75

Using the pwr package [@pwr]:

```{r}
library(pwr)
pwr.f2.test(f2 = 0.82, u = 2, sig.level = 0.01, power = 0.9)

```

From above, we know that n = v + u +1, so we need 24 + 2 + 1 = **27 subjects.**

### Example: power

Suppose there is an annual 5k running race in a city. We are interested if we can predict an individual's time to complete the race using two variables: total hours spent training the past three months and average pace per mile during training runs. We can perform a multiple regression to model race completion time as a function of training time and pace. We want to sample enough subjects to detect an effect when our two independent (explanatory) variables explains 45% of variation of the dependent (response) variable, race completion time. We will have 30 participating subjects. Assume the following:

-   hypothesized effect size of (f^2^ = .45 / (1-.45)), f^2^ = 0.82; R^2^ = .45

-   numerator degrees of freedom u = 2 (two predictors)

-   denominator degrees of freedom v = n − u − 1 = 30 - 2 - 1 = 27

-   significance level = 0.01

Using the pwr package [@pwr]:

```{r}
library(pwr)
pwr.f2.test(f2 = 0.82, u = 2, v = 27, sig.level = 0.01)

```

The power is about 0.95.

This can be done using the powertools package [@powertools], we use the 'random' argument to specify that our variables are random (as opposed to fixed), and the 'v' argument for a verbose output:

```{r}
library(powertools) 
mlrF.overall(N = 30, p = 2, Rsq = 0.45, power = NULL, random = TRUE, v = TRUE)
```

The power is about 0.98

## Partial F Tests

A partial F test evaluates whether adding one or more predictors to a regression model significantly improves its ability to explain variation in the response variable.\
It compares a reduced model (fewer predictors) with a full model (more predictors) and tests whether the additional predictors contribute significantly to ( R^2^ ).

For sample size estimation based on power, we need the following:

-   incremental effect size (f^2^ = R^2^~full~ - R^2^~reduced~) / (1 - R^2^~full~)

-   number of predictors in full model (sometimes called p)

-   number of predictors in reduced model (sometimes called q)

-   significance level of test

-   desired power

For power estimation based on sample size, we need the following:

-   incremental effect size (f^2^ = R^2^~full~ - R^2^~reduced~) / (1 - R^2^~full~)

-   number of predictors in full model (sometimes called p)

-   number of predictors in reduced model (sometimes called q)

-   significance level of test

### Example: sample size

Suppose there is an annual 5k running race in a city. We are interested if we can predict an individual's completion time using a statitical model. We already have a model fit with two variables (training hours and average pace) that explains 35% of variation in race completion time. We now want to know how many participants we need to detect whether adding a third variable, **weekly distance**, improves model performance by an additional 10% of explained variance (so that ( R^2^~full~ = 0.45 )).

Assume the following:

-   R^2^~full~ = 0.45
-   R^2^~reduced~ = 0.35
-   p = 3 (predictors in full model)
-   q = 2 (predictors in reduced model)
-   significance level ( = 0.05 )
-   desired power ( = 0.80 )

Using the `powertools` package [@powertools]:

```{r}
library(powertools)
mlrF.partial(N = NULL, p = 3, q = 2, Rsq.red = 0.35, Rsq.full = 0.45, power = 0.80)
```

### Example: power

Suppose there is an annual 5k running race in a city. We are interested if we can predict an individual's completion time using a statitical model. We already have a model fit with two variables (training hours and average pace) that explains 35% of variation in race completion time. If we sample 30 participants, we now want to know how much power we have to detect whether adding a third variable , **weekly distance**, improves model performance by an additional 10% of explained variance (so that ( R^2^~full~ = 0.45 )).

Assume the following:

-   N = 30
-   R^2^~full~ = 0.45
-   R^2^~reduced~ = 0.35
-   p = 3 (predictors in full model)
-   q = 2 (predictors in reduced model)
-   significance level ( = 0.05 )

Using the `powertools` package [@powertools]:

```{r}
library(powertools)
mlrF.partial(N = 30, p = 3, q = 2, Rsq.red = 0.35, Rsq.full = 0.45, power = NULL)
```

## Logistic Regression

## Poisson Regression
