---
title: "T tests"
---

## Two sample t-test

For sample size estimation based on _power_, we need the following:

- true difference in means (effect size)
- population standard deviation
- power of test 
- significance level of test
- direction of test: one or two-sided

### Example

We wish to run an experiment to test if the mean price of what male and female students pay at a library coffee shop is different. Our null hypothesis is no difference. We want to sample enough students to detect a difference of at least 75 cents. We're not sure which group pays more, so we'll do a two-sided test. Assume the following: 

- true difference is 0.75 cents
- population standard deviation is $2.25.
- we desire 0.9 power
- our significance level will be 0.05
- this will be two-sided test

Using base R [@R]:

```{r}
power.t.test(delta = 0.75, sd = 2.25, sig.level = 0.05, power = 0.9,
             alternative = "two.sided")
```

We should plan to observe 191 in each group.

Using the pwr package [@pwr], we must express effect size as Cohen's d: difference between the means divided by the population (or pooled) standard deviation.

```{r}
library(pwr)
pwr.t.test(d = 0.75/2.25, sig.level = 0.05, power = 0.9, 
           alternative = "two.sided")
```

